---
title: "Spatial patterns of fire severity and bark beetle outbreak in the Cameron Peak Fire"
author: "Sarah Hart"
contact: "sarah.hart@colostate.edu"
date: "2026-05-11"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: FALSE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
link-citations: true
csl: global-change-biology.csl
bibliography: references.bib
---

```{r setup, echo=T, results=T, message=F, warning=TRUE, progress=FALSE}
knitr::opts_chunk$set(
	echo = T,
	fig.align = "center",
	fig.height = 4,
	fig.width = 7,
	out.width= "504pt",
	message = FALSE,
	warning = FALSE,
	dpi = 300,
	progress = FALSE,
	cache = TRUE
)

### Import libraries
if (!require("remotes"))  install.packages("remotes")
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  here, # easy file structure
  tidyverse, # data manipulation
  ggplot2, # figures
  RColorBrewer, # color palettes
  patchwork, # combining figures
  sf, # spatial data
  terra # raster data 
) 


# Set seed
set.seed(2020)

# Set custom plotting theme
theme_new <- function(base_size = 9,base_family = "Helvetica"){
  theme_classic(base_size = base_size, base_family = base_family) %+replace%
    theme(
      axis.line.x = element_line(color="black", linewidth = 0.25),
      axis.line.y = element_line(color="black", linewidth = 0.25),
      axis.title = element_text(size = 9),
      axis.text = element_text(colour="black", size=8),
      legend.key=element_rect(colour=NA, fill =NA),
      panel.grid = element_blank(),   
      plot.background = element_rect(fill = NA, colour = NA),
      panel.border = element_rect(fill = NA, colour = NA),
      panel.background = element_rect(fill = "white", colour = "black"), 
      strip.background = element_rect(fill = "white"),
      strip.text = element_text(size = 9)
      
    )
}
theme_set(theme_new())
options(ggplot2.continuous.colour="viridis")
terraOptions(progress=0)
```

```{r create directories if they do not exist}
dir.create(here("Spatial/"), showWarnings = FALSE)
dir.create(here("Spatial/Raw"), showWarnings = FALSE)
dir.create(here("Spatial/Processed"), showWarnings = FALSE)
```

# Overview

This code compares the patterns of overlap between mortality of lodgepole pine (*Pinus contorta*) attributed to the mountain pine beetle (*Dendroctonus ponderosae)* and subsequent fire severity in the Cameron Peak Fire, which burned in 2020 in northwestern Colorado.

# Data

## Basic Geographic data

We obtained spatial data describing state boundaries from the [@u.s.censusbureau2019].

```{r DL_States}
## Download states (if not already downloaded)
if(!dir.exists(here("Data", "Spatial", "Raw", "States"))){
  dir.create(here("Data", "Spatial", "Raw", "States"), showWarnings = FALSE)
  temp <- here("Data", "Spatial", "Raw", "States", "cb_2018_us_state_20m.zip")
  download.file("https://www2.census.gov/geo/tiger/GENZ2018/shp/cb_2018_us_state_20m.zip", temp, mode="wb")
  unzip(temp, exdir=here("Data", "Spatial", "Raw", "States"))
  file.remove(temp)
}
```

```{r DL_ecoregions}
# Download Ecoregion data
dir.create(here("Data", "Spatial", "Raw", "Ecoregions"), showWarnings = FALSE)
temp <- here("Data", "Spatial", "Raw", "Ecoregions","us_eco_l3.zip")
if(!file.exists(temp)){
download.file("https://dmap-prod-oms-edc.s3.us-east-1.amazonaws.com/ORD/Ecoregions/us/us_eco_l3.zip", temp, mode="wb")
  unzip(temp, exdir=here("Data", "Spatial","Raw",  "Ecoregions"))
}
```

## Wildfire

### Wildfire perimeters

We acquired polygon data describing the extent of the area burned in the Cameron Peak fire from the Monitoring Trends in Burn Severity (MTBS) Project [-@mtbsproject2024].

```{r DL_MTBS}
## MTBS perimeters (if not already downloaded)
# Note downloading here from the USFS FSGeodata Clearinghouse because the MTBS website kept throwing errors when downloading data programmatically
if(!dir.exists(here("Data", "Spatial", "Raw", "Fire", "MTBS"))){
  dir.create(here("Data", "Spatial", "Raw", "Fire"), showWarnings = FALSE)
  dir.create(here("Data", "Spatial", "Raw", "Fire", "MTBS"), showWarnings = FALSE)
  temp <- here("Data", "Spatial", "Raw", "Fire", "MTBS", "mtbs_perimeter_data.zip")
  download.file("https://data.fs.usda.gov/geodata/edw/edw_resources/shp/S_USA.MTBS_BURN_AREA_BOUNDARY.zip", temp, mode="wb")
  unzip(temp, exdir=here("Data", "Spatial", "Raw", "Fire","MTBS"))
 file.remove(temp)
}
```

### Fire severity

As an estimate of fire severity we calculated the composite burn index (CBI) from Landsat multispectral imagery in Google Earth Engine using methods outlined by Parks et al. [-@parks2019]. Briefly CBI is modeled as function of four remotely sensed indices, the relativized burn ratio (RBR) [@parks2014], delta mid-infrared bi-spectral index (dMIRBI) [@mccarley2018], post-fire mid-infrared bi-spectral index (pMIRIBI) [\@mccarley2018], and delta normalized differenced vegetation index (dNDVI) [@pettorelli2005] and climatic water deficit (CWD), and latitude. RBR is calculated as follows:

$$ RBR =  \frac{dNBR}{NBR_{prefire}-0.001} $$

where NBR is the difference between near infrared (NIR) band and shortwave infrared (SWIR1) band divided by their sum and dNBR is the difference between prefire and postfire NBR. The mid-infrared bi-spectral index (MIRBI) is calculated from the SWIR1 and SWIR2:

$$ MIRBI = (10\times \rho_{SWIR1}) - (9.8\times\rho_{SWIR2})+2$$

$$dMIRBI = (MIRBI_{prefire} - MIRBI_{postfire})$$

and dMIRBI is the difference between prefire and postfire MIRBI. Finally, dNDVI is the difference between the pre- and post-fire normalized differenced vegetation index (NDVI), which is calculated as the difference between the NIR and red bands divided by their sum. Because the model over predicts CBI at low values, so we used Parks et al.'s [-@parks2019] bias correction, which is calculated as follows:

$$CBI_{bc} = \begin{cases} \ (CBI_{pred}-1.5)\times 1.3) + 1.5, & \ CBI_{pred}≤1.5\\ \ (CBI_{pred}-1.5)\times 1.715) + 1.5, & \ CBI_{pred}>1.5 \end{cases}$$

We then classified the resulting CBI scores into four burn severity levels, unburned (CBI: 0–0.1), low severity (CBI: 0.1–1.24), moderate severity (CBI: 1.25–2.24), and high severity (CBI: 2.25–3), following Key and Benson [-@key1999].

```{r prepGEECBI}
## Prepares fires perimeter dataset for use in Sean Parks' GEE code for calculating fire severity
dir.create(here("Data", "Spatial", "Processed", "GEE"))
dir.create(here("Data", "Spatial", "Processed", "GEE", "FirePerimeters"))
st_read(here('Data/Spatial/Raw/Fire/MTBS/S_USA.MTBS_BURN_AREA_BOUNDARY.shp'), quiet = TRUE) %>% 
  filter(YEAR==2020) %>% 
  filter(FIRE_NAME=="CAMERON PEAK") %>% 
  mutate(date=as.Date(as.character(IG_DATE), format="%Y%m%d")) %>% # conver ignition date to date object
  mutate(Start_Day=lubridate::yday(date)-14) %>% # set first possible date 
  mutate(End_Day=max(lubridate::yday(date))+30) %>% # set last possible date 
  mutate(Fire_Year=YEAR, Fire_ID=FIRE_ID) %>% # create variables with column names used in Parks code
  select(Fire_ID, Fire_Year, Start_Day, End_Day) %>% # include only Fire ID, Fire Year, Start Day and End Day columns in dataset
  mutate(Start_Day=ifelse(Start_Day<1, 1, Start_Day), End_Day=ifelse(End_Day>365, 365, End_Day)) %>% 
  st_write(here("Data", "Spatial", "Processed", "GEE", "FirePerimeters", paste0("CameronPeak.shp")), delete_dsn = TRUE, quiet = TRUE)
```

## Bark beetle outbreaks

We acquired 30 x 30 m raster data describing the cumulative extent of bark beetle-attributed tree mortality for the 1997-2019 period produced by Rodman et al. [-@rodman2021EffectsBarkBeetle]. Because Rodman et al.'s [-@rodman2021EffectsBarkBeetle] data does not distinguish among damaged caused by the MPB to lodgepole pine, the spruce beetle (*Dendroctonus rufipennis*) to Englemann spruce (*Picea engelmannii),* or the western balsam bark beetle (*Dryocoetes confusus*) to subalpine fir (*Abies lasiocarpa*), we obtained data on the distribution of lodgepole pine prior to outbreak. Specifically, a obtained 240 x 240 m map of the basal area of lodgepole pine in ca. 2002 from the National Insect and Disease Risk Mapping project [@usfs2019NationalIndividualTree]. We then listed lodgepole as present in any cell where the basal area of lodgepole pine was least 10 m^2^/ha and masked the maps of bark beetle outbreak to only areas where lodgepole was present prior to outbreak [@chapman2012SpatiotemporalPatternsMountain; @tutland2023OverlappingOutbreaksMultiple].

# Analyses

We used a simple overlay analyses to characterize patterns of overlap between fire severity and prior mountain pine beetle outbreak across the Cameron Peak burn scar. Specifically, overlaid maps of classified fire severity with maps of the presence or absence of MPB outbreak in lodgepole pine forests and generated a contingency table comparing the frequency of overlap.

```{r overlay}
terraOptions(progress=0)
dir.create(here("Results/Figures"))
# Import extent of Cameron Peak fire
CameronPeak <- st_read(here("Data", "Spatial", "Processed", "Fire", "GEE", "FirePerimeters", paste0("CameronPeak.shp")), quiet=T)

# Import lodgepole 
lpp <- rast(here("Data/Spatial/Raw/ITSPM/b108.tif"))
lpp[lpp<10] <- NA
lpp[lpp>0]<- 1
dir.create(here("Data", "Spatial", "Processed", "ITSPM"), showWarnings = FALSE)
writeRaster(filename=here("Data/Spatial/Processed/ITSPM/lpp.tif"), lpp, overwrite=T)
  
# Convert lodgepole pine raster to polygon 
lpp.poly <- lpp %>% 
  crop(st_transform(CameronPeak, st_crs(lpp))) %>% 
  mask(st_transform(CameronPeak, st_crs(lpp))) %>% 
  as.polygons()  %>% 
  st_as_sf() %>% 
  filter(b108==1)
  
lpp.poly %>%  st_write(here("Data", "Spatial", "Processed", "ITSPM", "lpp_poly.gpkg"), delete_dsn=T, quiet = TRUE)

# Convert area affected by bark beetle to raster
bb <- rast(here("Data/Spatial/Raw/BarkBeetle/doi_10_5061_dryad_1rn8pk0sn__v20210307/Rodman et al. RS_DataArchive/Data/SpatialData/BBPresence_finalMasked.tif"))
bb.poly <- bb %>% 
  crop(st_transform(CameronPeak, st_crs(bb))) %>% 
  mask(st_transform(CameronPeak, st_crs(bb))) %>% 
  mask(st_transform(lpp.poly, st_crs(bb))) %>% 
  as.polygons()  %>% 
  st_as_sf() %>%   filter(BBPresence_finalMasked==1)

 bb.poly %>% st_write(here("Data", "Spatial", "Processed", "bb_poly.gpkg"), delete_dsn=T, quiet = TRUE)

lpp.poly.affected <- lpp.poly %>% st_intersection(bb.poly) %>% st_make_valid() 
lpp.poly.affected %>%  st_write(here("Data", "Spatial", "Processed", "ITSPM", "lpp_poly1.gpkg"), delete_dsn=T, quiet = TRUE)
 
lpp.poly.unaffected <- lpp.poly %>% st_difference(bb.poly) %>% st_make_valid() 
lpp.poly.unaffected %>%  st_write(here("Data", "Spatial", "Processed", "ITSPM", "lpp_poly0.gpkg"), delete_dsn=T, quiet = TRUE)


# Import fire severity

cp <- rast(here("Data", "Spatial", "Processed", "Fire", "GEE", "CBI", "CO4060910587920200813_CBI_bc.tif"))

reclass_matrix <- matrix(c(
    0, 0.1, 0,
    0.1, 1.25, 1,
    1.25, 2.25, 2,
    2.25, 3, 3), ncol=3, byrow=TRUE)

  cp.class <- classify(cp, reclass_matrix, right=F)
  
cp.class1 <- cp.class %>% mask(st_transform(lpp.poly.affected, st_crs(cp.class)))
cp.class1 %>% writeRaster(here("Data", "Spatial", "Processed", "cp_class1.tif"), overwrite=T)
cp.class0 <- cp.class %>% mask(st_transform(lpp.poly.unaffected, st_crs(cp.class)))
cp.class0 %>% writeRaster(here("Data", "Spatial", "Processed", "cp_class0.tif"), overwrite=T)
  
  x <- cp.class1 %>% st_drop_geometry() %>% as.data.frame() %>% mutate(variable="affected")
  y <- cp.class0 %>% st_drop_geometry() %>% as.data.frame() %>% mutate(variable="unaffected")
  xy <- bind_rows(x,y) %>% filter(CBI_bc>0) %>% group_by(variable, CBI_bc) %>% summarise(n=n()) %>% group_by(variable) %>% mutate(prop=n/sum(n)) %>% unite(col="fill", variable:CBI_bc, sep="_", remove=F) %>% mutate(CBI_bc=factor(CBI_bc, levels=c(1,2,3), labels=c("low", "moderate", "high")))
  p1 <- xy %>% ggplot(aes(x=CBI_bc, y=prop, fill=fill)) + geom_bar(stat="identity", position="dodge") + scale_fill_manual(values=c("affected_1"="#1f78b4", "affected_2"="#ff7f00", "affected_3"="#e31a1c", "unaffected_1"="#a6cee3", "unaffected_2"="#fdbf6f", "unaffected_3"="#fb9a99"), labels=c("MPB & low CBI", "MPB & mod. CBI", "MPB & high CBI", "No MPB & low CBI", "No MPB & mod. CBI", "No MPB & high CBI")) + labs(x="composite burn index", y="Proportion of pixels") + theme(legend.title=element_blank(), legend.position=c(0.3, 0.82))+ theme(legend.text=element_text(size=7), axis.title=element_text(size=8), axis.text = element_text(size=7))+
  theme(legend.key.size = unit(0.25, 'cm')) 
  ggsave(p1, filename=here("Results/Figures/fig1.jpg"), width=2.5, height=2.55, units="in", dpi=900)

```

# References
